---
title: "Course Project 1 Repro_Research"
author: "Santiago Humberto Londoño Restrepo"
date: "13 de septiembre de 2016"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


#Loading and preprocesing the data

Load date into bd  and transform interval variable in appropiated format, it´s
transformed in date class.
```{r}
bd <- read.csv("data/activity.csv", stringsAsFactors = FALSE)

bd$date <- as.Date(bd$date)
```

You need to load the following packages:
```{r}
library(dplyr)
library(ggplot2)
```

#What is mean total number of step taken per day?

Calculate the total number of steps taken per day into bd1
```{r}
#library(dplyr) For calculate the total number of steps by day

bd1 <- group_by(bd, date) %>% summarize(total_numbers_steps=sum(steps))
bd1 <- as.data.frame(bd1)
bd1
```

Histogram of the total number of steps taken each day
```{r}
#library(ggplot2) My favorite plot system

ggplot(bd1, aes(total_numbers_steps)) + geom_histogram(bins = 9, fill="white",
                                                       colour="black") +
        ggtitle("Total number of steps taken each day") +
        ylab("Frecuency") + xlab("Number of steps")
```

Mean and median of the total number of steps taken per day
```{r}
#You need load dplyr package

bd2 <- summarize(bd1,mean_total = mean(total_numbers_steps, na.rm = TRUE), median_total = median(total_numbers_steps, na.rm = TRUE))
bd2 <- as.data.frame(bd2)
mean <- bd2[1]
median <- bd2[2]
```

The mean and median of the total number of steps taken per day is **`r mean`** and **`r median`** respectively.

#What is the average daily activity pattern?

Time series plot of the 5-minute interval and the average number of steps taken, averaged across all days

```{r}
#You need load dplyr and ggplot2 packages

bd3 <- group_by(bd, interval) %>% summarize(mean_steps_interval = mean(steps, na.rm = TRUE))
ggplot(bd3, aes(x=interval, y=mean_steps_interval)) + geom_line() +
        ggtitle("Average number of steps taken per 5-minute interval") +
        ylab("Average number of steps") + xlab("5-minute interval")
        
```

Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?

```{r}
hig_num_steps<- bd3[which.max(bd3$mean_steps_interval),]
```

**`r hig_num_steps[1]`** is 5-minute interval with the maximum average numbers of steps, it's iqual to  **`r hig_num_steps[2]`**



#Imputing missing values

Calculate and report the total number of missing values in the dataset (i.e. the total number of rows with NAs)
```{r}
totNA <- sum(is.na(bd$steps))
```

The total number of missin values is **`r totNA`**

Devise a strategy for filling in all of the missing values in the dataset bd. The strategy was fill NA values with median number of steps taken per 5-minute interval because the mean is far from median. To achieve this objective, I have created fillNA function that allow filling missing values in bd data set and creating a new dataset that is equal to the original dataset but with the missing data filled in.

```{r}
#You need load dplyr and ggplot2 packages

#bd4 <- group_by(bd, interval) %>% summarize(mean_interval = mean(steps, na.rm=TRUE), median_interval = median(steps, na.rm = TRUE))

#Below plot show that there is a huge difference between mean and median, so
# in this case is better choose median.
#ggplot(bd4, aes(interval)) + geom_line(aes(y=mean_interval, colour="mean_interval")) + geom_line(aes(y=median_interval, colour="median_interval"))

#fillNA function allow fill missing values in bd data set. bd4 data set is so important to fillNA performance
fillNA <- function(dataset){
        bd4 <- group_by(bd, interval) %>% summarize(mean_interval = mean(steps, na.rm=TRUE), median_interval = median(steps, na.rm = TRUE))
        
        bd4 <- as.data.frame(bd4)
        
        for (i in 1:dim(dataset)[1]) {
        if(is.na(dataset$steps[i])){
                dataset$steps[i] <- bd4[bd4$interval==dataset$interval[i],3]
        }
        }
        dataset
}

#Copy the data without NA
bdfill <- fillNA(bd) 

```


Histogram of the total number of steps taken each day with dataset without NA´s


```{r}
#You need load dplyr and ggplot2 packages
bd5 <- group_by(bdfill, date) %>% summarize(total_numbers_steps = sum(steps))

ggplot(bd5, aes(total_numbers_steps)) + geom_histogram(bins = 9, fill="white",
                                                       colour="black") +
        ggtitle("Total number of steps taken each day") +
        ylab("Frecuency") + xlab("Number of steps")
        
```

Mean and median total number of steps taken per day.

```{r}
bd6 <- summarize(bd5, mean_total = mean(total_numbers_steps),
                 median_total = median(total_numbers_steps))

       
meanFill <- bd6[1]
medianFill <- bd6[2]
```

The mean and median of the total number of steps taken per day with dataset without NA´s is **`r meanFill`** and **`r medianFill`** respectively. There is difference betweeen first and second part, specifically in the average. For the first part mean was `r mean`, more than a thousand step difference, this reflect the importance of 0 steps in the distribuction.



#Are there differences in activity patterns between weekdays and weekends?

Create a new factor variable in the dataset with two levels - "weekday" and "weekend" indicating whether a given date is a weekday or weekend day.
```{r}

newFactor <- function(dataset){
weekdays <- c("lunes", "martes", "miércoles", "jueves", "viernes")
weekends <- c("sábado", "domingo")
dataset$week1 <- weekdays(dataset$date)

for (j in 1:dim(dataset)[1]) {
        if((dataset$week1[j] %in% weekdays) | dataset$week1[j] == "miércoles"){
                dataset$week[j] <- "weekday"
        }else{
                dataset$week[j] <- "weekend"
        }
}
dataset
} #This function create new dataset with factor variable requested.

bd7 <- newFactor(bd)
        
```

Make a panel plot containing a time series plot of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis). 

```{r}
#You need load dplyr and ggplot2 packages
bd8 <- group_by(bd7, interval, week) %>% summarize(mean_steps_interval = mean(steps, na.rm = TRUE))
ggplot(bd8, aes(x=interval, y=mean_steps_interval)) + geom_line() +
        facet_grid(week ~ .) +
        ggtitle("Average number of steps taken per 5-minute interval and per week") +
        ylab("Average number of steps") + xlab("5-minute interval")

```